Automatic Matching of Bullet Lands

### 2016-01-20 imaging-paper.Rnw
proof-reading main text done

\begin{enumerate}
    \item At a fixed height $x$ extract a bullet's profile (Figure~\ref{fig:loess_step1}, with $x = 243.75$\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \emph{smoothing factor} $s$. (Figure~\ref{fig:loess_step3}, smoothing factor $s = 35$ corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first doubly-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, where $i$ is between 1  and $\lfloor s/2 \rfloor$). We call the location of this peak $p_{\ell}$ (see Figure~\ref{fig:loess_step47}). 
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{\ell}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{r}, v_{r})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{\ell}$ and $y_i > v_{r}$) (see Figure~\ref{fig:loess_step47}).
\end{enumerate}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures at cross sections 25$\mu m$ apart. For a stable region, we require a minimum of 0.9 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and 
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are running $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
 When things go perfectly, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 42 pairs of lands with at least thirteen CMS in common are matches. 
%
\begin{figure}[hbtp]
  \centering
%' \begin{subfigure}[b]{\textwidth}\centering
%' \caption{}
\begin{minipage}[t]{.47\textwidth}
<<cms-bars, echo=FALSE, fig.width=7, fig.height=3.5, out.width='\\textwidth'>>=
ggplot(data=subset(bstats, !flagged)) + 
  geom_bar(aes(x=factor(CMS))) + theme_bw() +
  theme(legend.position="bottom") + 
  xlab("maximum CMS")
@
%\end{subfigure}
%\begin{subfigure}[b]{\textwidth}\centering
\end{minipage}
\begin{minipage}[t]{.52\textwidth}
<<cms-spines, echo=FALSE,  fig.width=7, fig.height=4, out.width='\\textwidth'>>=
bstats$km <- c( "KNM", "KM")[as.numeric(bstats$match)+1]
ggplot(data=subset(bstats, !flagged)) + 
  geom_bar(aes(x=factor(CMS), fill=km), position="fill") +
  theme(legend.position="bottom") + 
  scale_fill_brewer("", palette="Paired") +
  xlab("maximum CMS") + ylab("Proportion")
@
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$CMS[!bstats$match], na.rm=T)} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$CMS[!bstats$match])}). Known matches share on average \Sexpr{mean(bstats$CMS[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$CMS[bstats$match])}. While the probability for a match increases with the number of maximal CMS, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch, between two lands that share twelve consecutively matched striae. Visually we can easily tell that these two lands do not match well.

@@

print(ggplot() +
  theme_bw() + 
  geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), ymin=-6, ymax=5, data=res$lines, alpha=0.2, show.legend=FALSE) +
  geom_line(aes(x = y, y = l30, linetype = bullet),  data = res$bullets) +
  scale_linetype_discrete("") + ylab("") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey50", alpha=0.5)) +
  theme(legend.position = c(1,1), legend.justification=c(1,1)) + 
  ylim(c(-6,6)) +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(res$lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(res$lines, match)))
@
\caption{\label{fig:mismatch}Known mismatch with a relatively large number of maximal consecutive matching striae (twelve) in the middle. The pattern in the middle does look surprisingly similar, however the outer ends of the signatures easily reveals this as mismatch. }
\end{figure}

@@

Another benefit of the digitized version of the images is that we can apply several hundred decision trees to combine in a random forest~\citep{breiman:2001, randomForest}.  For each of the trees in a random forest, only two thirds of the observations are used for fitting, while  the remaining third is used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=6, fig.height=3, out.width='.6\\textwidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, even for the single tree the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants, who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, resulting in a false negative rate of 0.0011.

However, error rates based on bullet-to-bullet matches do not carry a lot of weight because of the small size of the study: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger experiment. 
Another thing to note about the  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on 0.5 is not necessary. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we could therefore introduce an interval of `inconclusive' results in the middle of the spectrum -- which turns out to be not necessary in the Hamby study, because, here, the results from the random forest are very clear cut. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree and the random forest. %\hh{The tree predicts 132 known matches correctly, with 40 false negatives and one false positive match. The random forest predicts all 172 known matches correctly as well as all 10,212 known non-matches.} 
%Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

@@

\hh{Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement section~\ref{supp:extended}.}

@@

The matching algorithm is sensitive to the parameter choices made. The distance between cross sections (currently 25$\mu m$) to evaluate stability, as well as the cross-correlation factor (currently 0.9) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). While we tried to lay out in the paper the impact that each of the parameter choices has on the matching performance, we are still far from an optimized scenario.

@@

\hh{XXX speed? XXX}

>>>>

\begin{enumerate}
    \item At a fixed height $x$ extract a bullet's profile (Figure~\ref{fig:loess_step1}, with $x = 243.75\mu m$).
    \item For each $y$ value, smooth out any deviations occurring near the minima by twice applying a rolling average with a pre-set \emph{smoothing factor} $s$. (Figure~\ref{fig:loess_step3}, smoothing factor $s = 35$ corresponding to 55$\mu m$).
%    \item For each smoothed $y$ value, compute another rolling average using the same smoothing factor as above. (Figure~\ref{fig:loess_step3}).
    \item Determine the location of the peak of the left groove by finding the first doubly-smoothed value $y_i$ that is the maximum within its smoothing window (e.g.\ such that $y_i > y_{i - 1}$ and $y_i > y_{i + 1}$, where $i$ is between 1  and $\lfloor s/2 \rfloor$). We call the location of this peak $p_{\ell}$ (see Figure~\ref{fig:loess_step47}). 
    \item Similarly, determine the location of the valley of the left groove by finding the first double-smoothed $y_j$ that is the minimum within its smoothing window. Call the location of this valley $v_{\ell}$.
    \item Reverse the order of the $y$ values and repeat the previous two steps to find the peak and valley of the right groove, $(p_{r}, v_{r})$.
    \item Trim the surface measurements to values within the two grooves (i.e.\ remove all records with $y_i < v_{\ell}$ and $y_i > v_{r}$) (see Figure~\ref{fig:loess_step47}).
\end{enumerate}

@@

\section{Evaluation}
In order to get a better understanding of  how  the matching algorithm works in known matches and non-matches, we investigate its performance in a test scenario based on the James Hamby study.
As a first step, we automatically assess the quality of each of the lands by  checking that we can identify a stable region on each land. For this, we compute the cross-correlation of signatures extracted from heights 25$\mu m$ apart. For a stable region, we require a minimum of 0.95 for the cross correlation. Four lands from different bullets are flagged as problematic in this respect. A visual inspection (see Figure~\ref{fig:fourflags}) shows that each one of these lands has scratch marks across the surface, also known as `tank rash'.
%
\begin{figure}
  \centering
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 6 Bullet 2-1}
%' <<echo=FALSE>>=
%' b1 <- read.x3p(file.path(datadir, "Br6 Bullet 2-1.x3p"))
%' image(b1$surface.matrix)
%' @
\includegraphics[width=\textwidth]{images/br6-2-1-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Barrel 9 Bullet 2-4}
\includegraphics[width=\textwidth]{images/br9-2-4-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet B-2}
\includegraphics[width=\textwidth]{images/b-2-grey.png}
\end{subfigure}
\begin{subfigure}[t]{.49\textwidth}\centering
\caption{Unknown Bullet Q-4}
\includegraphics[width=\textwidth]{images/q-4-grey.png}
\end{subfigure}
\caption{\label{fig:fourflags}Images of the four lands that got flagged during the quality assessment. All of them show scratch marks (tank rash) across the striation marks from the barrel. They are excluded from the remainder of the analysis.}
\end{figure}
%
We exclude these four lands from further matching considerations and
 run all remaining lands from the unknown bullets against all remaining lands from known bullets for matches, i.e. we are comparing $15 \times 6 -2 = 90 - 2 = 88$ lands from unknown bullets against $2 \times 10 \times 6 -2 = 120 - 2 = 118$ lands from known bullets, yielding a total of $10,384$ land-to-land comparisons. Out of these comparisons, there are 172 known matches (KM), while the rest are known non-matches (KNM). 
 When things go well, they look like the results in Figure~\ref{fig:hamby-perfect}: Figure~\ref{fig:hamby-perfect}a shows the distribution of the number of maximum consecutive matching striae between land C-3 and all 118 lands from known bullets. Two lands show a high CMS. These correspond to the known matches with C-3, shown in Figures~\ref{fig:hamby-perfect}b and~\ref{fig:hamby-perfect}c. 
 Unfortunately, not all results are as clear cut. 
It might not be reasonable to assume that we can match all lands, but the idea is to try to maximize the number of matches to get an overview of what we might be able to expect from an automated match.

@@

Figure~\ref{fig:cms} shows the strong connection between the maximal number of consecutive striae and matches in the Hamby study. All 42 pairs of lands with at least thirteen CMS in common are matches. 
%
\begin{figure}[hbtp]
  \centering
%' \begin{subfigure}[b]{\textwidth}\centering
%' \caption{}
\begin{minipage}[t]{.47\textwidth}
<<cms-bars, echo=FALSE, fig.width=7, fig.height=3.5, out.width='\\textwidth'>>=
ggplot(data=subset(bstats, !flagged)) + 
  geom_bar(aes(x=factor(CMS))) + theme_bw() +
  theme(legend.position="bottom") + 
  xlab("maximum CMS")
@
%\end{subfigure}
%\begin{subfigure}[b]{\textwidth}\centering
\end{minipage}
\begin{minipage}[t]{.52\textwidth}
<<cms-spines, echo=FALSE,  fig.width=7, fig.height=4, out.width='\\textwidth'>>=
bstats$km <- c( "KNM", "KM")[as.numeric(bstats$match)+1]
ggplot(data=subset(bstats, !flagged)) + 
  geom_bar(aes(x=factor(CMS), fill=km), position="fill") +
  theme(legend.position="bottom") + 
  scale_fill_brewer("", palette="Paired") +
  xlab("maximum CMS") + ylab("Proportion")
@
\end{minipage}
%' \end{subfigure}
%' \begin{subfigure}[t]{\textwidth}\centering
%' \caption{\label{fig:cms-dens}Densities of maximum CMS for Known Matches (KM) and Known Non Matches (KNM).}
%' <<cms-densities, echo=FALSE,  fig.width=7, fig.height=3.5, out.width='.6\\textwidth'>>=
%' x <- seq(0,29, by=0.03)
%' km <- sm.density(bstats$CMS[bstats$match],h=2, model="Normal", display="none", eval.points=x)
%' knm <- sm.density(bstats$CMS[!bstats$match],h=1, display="none", eval.points=x)
%' dframe <- data.frame(x = x, km=km$estimate, knm=knm$estimate)
%' dframe$kmtail <- rev(cumsum(rev(dframe$km)))/sum(dframe$km)
%' dframe$knmtail <- rev(cumsum(rev(dframe$knm)))/sum(dframe$knm)
%' 
%' densities <- data.frame(
%'   match = rep(c("KM", "KNM"), each=(length(x)+2)),
%'   x = rep(c(0, x, 29), times = 2),
%'   estimate = c(0, km$estimate, 0, 0, knm$estimate,0))
%' 
%' ggplot() +
%'   geom_polygon(aes(x=x, y=estimate, fill=match, group=match), colour="black",
%'                data=densities, alpha=0.5) +
%' xlab("maximum CMS") +
%'   scale_fill_brewer("", palette="Paired") +
%'   ylab("Density estimates") +
%'   theme_bw() + 
%'   theme(legend.position=c(1,1), legend.justification=c(1,1),
%'         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' #  geom_line(aes(x=x, y=pmin(0.2, kmtail/knmtail/10)), data=dframe) 
%' 
%' # ggplot(data=bstats) + 
%' #     geom_density(aes(x=cms, fill=match), alpha=0.5) +
%' #   xlab("maximum CMS") +
%' #   scale_fill_brewer("", palette="Paired") +
%' #   theme_bw() + 
%' #   theme(legend.position=c(1,1), legend.justification=c(1,1),
%' #         legend.background = element_rect(fill=alpha('white', 0.4))) 
%' 
%' @
%' \end{subfigure}
\caption{\label{fig:cms}Distribution of maximal CMS (left). Conditional barchart \citep{hummel} on the right: heights show probability of match/non-match given a specific CMS. All land-to-land comparisons with at least 13 CMS are matches.}
\end{figure}
%
There are two things that should be noted at this point: the automated algorithm finds a relatively high number of CMS even for non-matches. On average, there are \Sexpr{mean(bstats$CMS[!bstats$match], na.rm=T)} maximal CMS between known non-matches (with a standard deviation of \Sexpr{sd(bstats$CMS[!bstats$match])}). Known matches share on average \Sexpr{mean(bstats$CMS[bstats$match])} maximal CMS, with a standard deviation of \Sexpr{sd(bstats$CMS[bstats$match])}. While the probability for a match increases with the number of maximal CMS, a large number of maximal CMS by itself is not indicative of a match, as was previously pointed out by \citet{miller:1998}. Figure~\ref{fig:mismatch} shows a known mismatch between two lands that share twelve consecutively matched striae. Visually we can easily tell that these two lands do not match well.

@@

print(ggplot() +
  theme_bw() + 
  geom_rect(aes(xmin=xmin, xmax=xmax, fill=factor(type)), ymin=-6, ymax=5, data=res$lines, alpha=0.2, show.legend=FALSE) +
  geom_line(aes(x = y, y = l30, linetype = bullet),  data = res$bullets) +
  scale_linetype_discrete("") + ylab("") +
  scale_fill_brewer("", palette="Set2", na.value=alpha("grey50", alpha=0.5)) +
  theme(legend.position = c(1,1), legend.justification=c(1,1)) + 
  ylim(c(-6,6)) +
  geom_text(aes(x = meany), y= -5.5, label= "x", data = subset(res$lines, !match)) +
  geom_text(aes(x = meany), y= -5.5, label= "o", data = subset(res$lines, match)))
@
\caption{\label{fig:mismatch}Known mismatch with a relatively large number of maximal consecutive matching striae (twelve) in the middle. The pattern in the middle does look surprisingly similar, however the outer ends of the signatures easily reveals this comparison as mismatch. }
\end{figure}

@@

Another benefit of the digitized version of the images is that we can apply several hundred decision trees to combine in a random forest~\citep{breiman:2001, randomForest}.  For each of the trees in a random forest, only two thirds of the observations are used for fitting, while  the remaining third is used to evaluate the tree's predictive power and accuracy, or its reverse, the error rate. Because errors are determined from the one third of held-back observations, this error rate is called the out-of bag (OOB) error. 
Figure~\ref{fig:oob} shows the cumulative out-of-bag error (OOB) rate for 300 trees. 
%
\begin{figure}[hbtp]
  \centering
<<oob, echo=FALSE, fig.width=6, fig.height=3, out.width='.7\\textwidth'>>=
qplot(id, OOB, geom="line", data=errors) + theme_bw() + ylim(c(0,NA)) + xlab("Number of trees") + ylab("Out of Bag Error (OOB)")
@
\caption{\label{fig:oob}Cumulative out-of-bag error rate of a random forest fit to predict land-to-land matches from image features.}
\end{figure}
%
After about 100 trees, the error rate of land-to-land comparisons stabilizes at 0.0039. This is a weighted average between false positive error rate of 0.0001 and an error rate of false negatives of 0.2267. This out-of-bag error rate is over-estimating the actual error in the Hamby study: here, the final random forest based on 300 trees is able to correctly predict all known  matches and non-matches (see Figure~\ref{fig:tree-forest}).
Note that this error rate is based on land-to-land comparisons and is much lower for bullet-to-bullet matches. In the case of the Hamby study, even for the single tree the overall error rate is zero, if we require for a bullet match that at least two of a bullet's lands are matched. % (see Figure~\ref{fig:treeresult}). 
%' \begin{figure}[hbtp]
%'   \centering
%' <<overall-tree, echo=FALSE, fig.width=7, fig.height=4, out.width='.5\\textwidth'>>=
%' bstats$bullet <- gsub("-[0-9]$", "", bstats$b2)
%' bullets <- bstats %>% filter(match, !flagged) %>% group_by(bullet) %>% summarize(
%'   n = n(),
%'   pred = sum(pred > 0.5),
%'   forest = sum(forest > 0.5)
%' )
%' library(reshape2)
%' bm <- melt(data=bullets, id.var=c("bullet", "n"))
%' 
%' bm$prediction <- c("Tree", "Random Forest")[as.numeric(bm$variable)]
%' qplot(data=bm, 100*value/n, reorder(bullet, value/n), size=I(2.5), 
%'       shape=prediction, colour=prediction, alpha=I(0.6)) + 
%'   theme_bw()  + ylab("") + 
%' #  scale_shape_manual(values=c(1,2)) +
%'   scale_x_continuous("Percent of correctly predicted land-to-land matches", 
%'                      breaks = c(0,25,50,75,100), limits=c(0,100)) +
%'   theme(legend.position = "bottom")
%' @
%' \caption{\label{fig:treeresult}Overall predictions for each bullet based on the decision tree (maximum possible number is 12, except for Q and B, where the maximum is 11). Both results for the tree and the random forest are shown. In the tree, at least \hh{40\% of all possible} lands of a bullet are  matched correctly, while the random forest achieves perfect results. Unknown bullet B is the hardest for the tree to match.}
%' \end{figure}
%' %
This makes the errors in the automated approach smaller than the human error in the Hamby study. Out of the 507 participants, who returned results, eight (out of $15 \times 507 = 7,605$) bullets were not matched conclusively, corresponding to a rate of 0.0011. 
%\hh{XXX calling this a false negative rate is problematic. FEs will be up in arms, because inconclusive doesn't necessarily mean something was missed, it could also mean that there was something wrong with the bullets. XXX }

However, error rates based on bullet-to-bullet matches do not carry a lot of weight because of the small size of the study: fifteen unknown bullets are successfully matched to two pairs of ten bullets. Matching bullets can only be tested realistically in a much bigger experiment. 
Another thing to note about  the random forest's error rates is that they are based on probability cutoffs of 0.5, i.e.\ whenever the predicted probability of a match exceeds 0.5, a match is declared. Basing this decision on 0.5 is not necessary. In practice, examiners are allowed a third option of `inconclusive'. On a probability spectrum of outcomes we could therefore introduce an interval of `inconclusive' results in the middle of the spectrum -- which turns out to be not necessary in the Hamby study, because, here, the results from the random forest are very clear cut. Figure~\ref{fig:tree-forest} shows a comparison of the predicted probabilities of a match by the tree and the random forest. %\hh{The tree predicts 132 known matches correctly, with 40 false negatives and one false positive match. The random forest predicts all 172 known matches correctly as well as all 10,212 known non-matches.} 
%Overall, there is a lot of agreement; out of 10,800 land-to-land comparisons, only four cases result in a different decision.

@@

Besides including results from known matches against known non-matches, we can increase the number of comparisons in the Hamby study to include all possible land-to-land comparisons. This effectively doubles the number of data points available. Comparisons not previously included in fitting the random forest can also be used as an additional source for assessing error rates.  Results for this and a more detailed discussion can be found in Supplement section~\ref{supp:extended}.

@@

The matching algorithm is sensitive to the parameter choices made. The heights at which signatures are extracted (currently 25$\mu m$ apart) to evaluate stability, as well as the cross-correlation factor (currently 0.95) we set as a minimum threshold do affect the final outcome. Another parameter that must be selected is the amount of smoothing when identifying peaks and grooves (currently, a window of 23.4375$\mu m$ is used, corresponding to a window of 7 values to the left and the right of an observation). While we tried to lay out in the paper the impact that each of the parameter choices has on the matching performance, we are still far from an optimized scenario.

@@

%\hh{XXX speed? XXX}
